Annotation-Based Empirical Performance Tuning Using Orio

### 2013-11-03 doc/papers/iccs09/paper.tex

To generate highly optimized code from a MATLAB prototype of the composed
BLAS operation, we follow a three-step approach, illustrated in
Figure~\ref{fig:process}.  This process is repeated when the MATLAB code changes
or the code must be tuned on a new architecture.
To begin, we have developed a compiler that
%
\begin{wrapfigure}{r}{3in}
\vspace{-.2in}
\centering
\includegraphics[width=3in]{figures/process.pdf}
\vspace{-.15in}
\caption{Code generation and tuning process.}
\label{fig:process}
\vspace{-.1in}
\end{wrapfigure}
%
converts a MATLAB script into simple C code~\cite{Siek}.  After generating
the C code from the high-level MATLAB prototype, we (optionally) use 
the source-to-source automatic parallelization tool Pluto~\cite{Pluto} %developed at Ohio State University
to optimize for coarse-grained parallelism and locality simultaneously. Using
the results of the Pluto analysis, we insert annotations into the C code,
which are then processed by our extensible annotation system Orio to generate
many tuned versions of the same operation using different optimization
parameters. Orio then performs an empirical search to select the best among
the multiple optimized code variants.

@@

Figure~\ref{fig:orio-example} shows an annotation example used by Orio to
empirically optimize VADD operation on Blue Gene/P. The annotations contain
performance hints that instruct Orio to perform memory alignment
optimization, loop unrolling, and multicore parallelization (using
OpenMP). In addition to these simple optimizations, Orio supports other
transformations such as loop blocking, loop permutation, scalar replacement,
array copy optimization, and some architecture-dependent optimizations. The
right-hand side of Figure~\ref{fig:orio-example} shows separate tuning
specifications used for building and running executable tests, including
performance parameter values, execution environment details, input variable
information, and the search algorithm. Orio also supports parallel search
when parallel resources are available. In this example, the parallel Orio
driver simultaneously executes 64 code variants in the same parallel job. 
%The
%commands used by Orio to submit a parallel job and to query its status are
%also specified in the tuning specifications. 
At present, users must create the tuning specifications manually for
each architecture. When Orio is used in conjunction with compiler
tools, such as the MATLAB compiler described in this paper, it should
eventually be possible to automatically generate the tuning
specifications.

@@

We evaluated our approach by running experiments on an Intel Xeon
workstation and the Blue Gene/P at Argonne. The Intel machine has dual
quad-core E5462 Xeon processors (8 cores total) running at 2.8 GHz
(1600 MHz FSB) with 2 GB RAM, running Ubuntu 8.04. Intel C compiler
(v10.1) was used with \texttt{-O3} option (and
\texttt{-parallel}/\texttt{-openmp} for automatic/manual
parallelization, respectively). Each node of the Blue Gene/P has four
850 MHz PowerPC 450 processors with a dual floating-point unit and 2
GB total memory per node, running a proprietary operating system. On the
Blue Gene/P, we used IBM XLC compiler (v9.0), with \texttt{-O3
-qstrict -qarch=450d -qtune=450 -qhot} options (and
\texttt{-qsmp=auto}/\texttt{-qsmp=noauto} for automatic/manual
parallelization, respectively).

@@

The performance results of tuning the VADD operation on the Blue
Gene/P are given in Figure~\ref{fig:results}(a). The ``Base'' label
designates the C implementation generated by the MATLAB compiler. We also
tested the performance of an implementation that calls
\texttt{DAXPY} twice using available BLAS libraries.
Finally, we tuned the simple C loop version using Orio, with the performance
annotations previously shown in Figure~\ref{fig:orio-example}. In this
experiment, we measured the performance for both the sequential and parallel
scenarios (indicated by (S) and (P) in the legend, respectively). Even for a very simple operation such as vector addition the
compiler alone is unable to obtain the same level of performance as the
empirically tuned versions. Furthermore, as expected, the BLAS implementation
does not exploit locality and thus performed worse than the single-loop
implementation.
%of the single loop implementation due to l
% however, selecting the correct compiler
%options was crucial for achieving good performance of the base version.

@@

The Xeon performance results of ATAX are shown in
Figure~\ref{fig:results}(b). The Orio-tuned version that incorporates
Pluto-generated loop fusion optimizations and Orio parallelization
directives achieves the best performance for most problem sizes,
outperforming the Intel MKL version by a factor of 2 to 5.7 and the
compiler-optimized C version by a factor of 4 to 7. The optimizations
performed by both Orio versions include scalar replacement,
vectorization, and loop unroll/jam.

@@

The performance for the BiCG kernel operation is shown in
Figure~\ref{fig:results}(d). For this operation, the Pluto analysis did not
result in performance improvement. Thus we are showing the results obtained
only through Orio transformations, which included vectorization, scalar
replacement, and loop unroll/jam. Again the best performance was achieved
by the parallel Orio version, while all the BLAS versions performed worse
than the compiler-optimized C loop version.

>>>>

To generate highly optimized code from a MATLAB prototype of the composed
BLAS operation, we follow a three-step approach, illustrated in
Fig.~\ref{fig:process}.  This process is repeated when the MATLAB code changes
or the code must be tuned on a new architecture.
To begin, we have developed a compiler that
%
\begin{wrapfigure}{r}{3in}
\vspace{-.2in}
\centering
\includegraphics[width=3in]{figures/process.pdf}
\vspace{-.15in}
\caption{Code generation and tuning process.}
\label{fig:process}
\vspace{-.1in}
\end{wrapfigure}
%
converts a MATLAB script into simple C code~\cite{Siek}.  After generating
the C code from the high-level MATLAB prototype, we (optionally) use 
the source-to-source automatic parallelization tool Pluto~\cite{Pluto} %developed at Ohio State University
to optimize for coarse-grained parallelism and locality simultaneously. Using
the results of the Pluto analysis, we insert annotations into the C code,
which are then processed by our extensible annotation system Orio to generate
many tuned versions of the same operation using different optimization
parameters. Orio then performs an empirical search to select the best among
the multiple optimized code variants.

@@

Figure~\ref{fig:orio-example} shows an annotation example used by Orio to
empirically optimize VADD operation on Blue Gene/P. The annotations contain
performance hints that instruct Orio to perform memory alignment
optimization, loop unrolling, and multicore parallelization (using
OpenMP). In addition to these simple optimizations, Orio supports other
transformations such as loop blocking, loop permutation, scalar replacement,
array copy optimization, and some architecture-dependent optimizations. The
right-hand side of Fig.~\ref{fig:orio-example} shows separate tuning
specifications used for building and running executable tests, including
performance parameter values, execution environment details, input variable
information, and the search algorithm. Orio also supports parallel search
when parallel resources are available. In this example, the parallel Orio
driver simultaneously executes 64 code variants in the same parallel job. 
%The
%commands used by Orio to submit a parallel job and to query its status are
%also specified in the tuning specifications. 
At present, users must create the tuning specifications manually for
each architecture. When Orio is used in conjunction with compiler
tools, such as the MATLAB compiler described in this paper, it should
eventually be possible to automatically generate the tuning
specifications.

@@

We evaluated our approach by running experiments on an Intel Xeon
workstation and the Blue Gene/P at Argonne. The Intel machine has dual
quad-core E5462 Xeon processors (8 cores total) running at 2.8 GHz
(1600 MHz FSB) with 16 GB RAM, running Ubuntu 8.04. Intel C compiler
(v10.1) was used with \texttt{-O3} option (and
\texttt{-parallel}/\texttt{-openmp} for automatic/manual
parallelization, respectively). Each node of the Blue Gene/P has four
850 MHz PowerPC 450 processors with a dual floating-point unit and 2
GB total memory per node, running a proprietary operating system. On the
Blue Gene/P, we used IBM XLC compiler (v9.0), with \texttt{-O3
-qstrict -qarch=450d -qtune=450 -qhot} options (and
\texttt{-qsmp=auto}/\texttt{-qsmp=noauto} for automatic/manual
parallelization, respectively).

@@

The performance results of tuning the VADD operation on the Blue
Gene/P are given in Fig.~\ref{fig:results}(a). The ``Base'' label
designates the C implementation generated by the MATLAB compiler. We also
tested the performance of an implementation that calls
\texttt{DAXPY} twice using available BLAS libraries.
Finally, we tuned the simple C loop version using Orio, with the performance
annotations previously shown in Fig.~\ref{fig:orio-example}. In this
experiment, we measured the performance for both the sequential and parallel
scenarios (indicated by (S) and (P) in the legend, respectively). Even for a very simple operation such as vector addition the
compiler alone is unable to obtain the same level of performance as the
empirically tuned versions. Furthermore, as expected, the BLAS implementation
does not exploit locality and thus performed worse than the single-loop
implementation.
%of the single loop implementation due to l
% however, selecting the correct compiler
%options was crucial for achieving good performance of the base version.

@@

The Xeon performance results of ATAX are shown in
Fig.~\ref{fig:results}(b). The Orio-tuned version that incorporates
Pluto-generated loop fusion optimizations and Orio parallelization
directives achieves the best performance for most problem sizes,
outperforming the Intel MKL version by a factor of 2 to 5.7 and the
compiler-optimized C version by a factor of 4 to 7. The optimizations
performed by both Orio versions include scalar replacement,
vectorization, and loop unroll/jam.

@@

The performance for the BiCG kernel operation is shown in
Fig.~\ref{fig:results}(d). For this operation, the Pluto analysis did not
result in performance improvement. Thus we are showing the results obtained
only through Orio transformations, which included vectorization, scalar
replacement, and loop unroll/jam. Again the best performance was achieved
by the parallel Orio version, while all the BLAS versions performed worse
than the compiler-optimized C loop version.

### 2013-11-03 doc/papers/ipdps09/abstract.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@427 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

In many scientific applications, significant time is spent in tuning codes for a particular high-performance architecture. Tuning approaches range from the relatively nonintrusive (e.g., by using compiler options) to extensive code modifications that attempt to exploit specific architecture features. Intrusive techniques often result in code changes that are not easily reversible, which can negatively impact readability, maintainability, and performance on different architectures. We introduce an extensible annotation-based empirical tuning system called Orio, which is aimed at improving both performance and productivity by enabling software developers to insert annotations in the form of structured comments into their source code that trigger a number of low-level performance optimizations on a specified code fragment.  To maximize the performance tuning opportunities, the annotation processing infrastructure is designed to support both architecture-independent and architecture-specific code optimizations. Given the annotated code as input, Orio generates many tuned versions of the same operation and empirically evaluates the alternatives to select the best performing version for production use. We have also enabled the use of the PLuTo automatic parallelization tool in conjunction with Orio to generate efficient OpenMP-based parallel code. We describe our experimental results involving a number of computational kernels, including dense array and sparse matrix operations.

>>>>

For many scientific applications, significant time is spent in tuning
codes for a particular high-performance architecture. Tuning
approaches range from the relatively nonintrusive (e.g., by using
compiler options) to extensive code modifications that attempt to
exploit specific architecture features. Intrusive techniques often
result in code changes that are not easily reversible, and can
negatively impact readability, maintainability, and performance on
different architectures. We introduce an extensible annotation-based
empirical tuning system called Orio that is aimed at improving both
performance and productivity. It allows software developers to insert
annotations in the form of structured comments into their source code
to trigger a number of low-level performance optimizations on a
specified code fragment.  To maximize the performance tuning
opportunities, the annotation processing infrastructure is designed to
support both architecture-independent and architecture-specific code
optimizations. Given the annotated code as input, Orio generates many
tuned versions of the same operation and empirically evaluates the
alternatives to select the best performing version for production
use. We have also enabled the use of the Pluto automatic
parallelization tool in conjunction with Orio to generate efficient
OpenMP-based parallel code. We describe our experimental results
involving a number of computational kernels, including dense array and
sparse matrix operations.
